---
title: "EvoNEST Data Science Pipeline"
author: "Interactive R Workflow"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This R Markdown notebook provides a step-by-step interface for inexperienced researchers to:

1.  **Fetch data** from the EvoNEST API
2.  **Process mechanical data** and fit polynomial models to stress-strain curves
3.  **Build analysis tables** from downloaded and processed data
4.  **Analyze outliers** using hierarchical sigma-based detection

**Instructions**: Simply run each code chunk in order by clicking the green "play" button (▶) or pressing `Ctrl+Shift+Enter`. The scripts will guide you through configuration when needed.

------------------------------------------------------------------------

# Section 1: Fetch EvoNEST Data

Run the chunk below to fetch samples, traits, and experiments from the EvoNEST API.

This will: - Prompt you to configure API credentials and fetch options - Download data to `downloaded_data/` folder

```{r fetch-data, eval=FALSE}
# Source and run the data fetch script
source("data_fetch.R")
```

------------------------------------------------------------------------

# Section 2: Process Mechanical Data

Run the chunk below to process mechanical test data and fit polynomial models to stress-strain curves.

This will: - Prompt you to configure fracture detection and polynomial fitting parameters - Process all experiments in `downloaded_data/experiments_data.json` - Generate polynomial coefficients and save results to `processed_data/hierarchical_experiment_data_no_curves.json` - Optionally save plots showing fitted curves

```{r process-mechanical, eval=FALSE}
# Source and run the mechanical data processing script
source("process_mechanical_data.R")
```

------------------------------------------------------------------------

# Section 3: Build Data Analysis Tables

Run the chunk below to load downloaded and processed data into tibbles for analysis.

This will create three tables: - **samples_df**: Sample metadata (animals, silk samples, taxonomy, location) - **traits_df**: Trait measurements (diameter, mechanical properties, morphology) - **experiments_df**: Processed experiment results with polynomial fits

The script will display summary statistics for each table.

```{r build-tables}
# Source the analysis script (this creates the tables)
result <- source("analyse_data.R", local = TRUE)

# The script returns the tables invisibly, but we can also access them
# from the environment since they're created in the global environment
cat("\n✅ Tibbles are now available for analysis:\n")
cat(sprintf("   - samples_df: %d rows × %d columns\n",
            nrow(samples_df), ncol(samples_df)))
cat(sprintf("   - traits_df: %d rows × %d columns\n",
            nrow(traits_df), ncol(traits_df)))
cat(sprintf("   - experiments_df: %d rows × %d columns\n",
            nrow(experiments_df), ncol(experiments_df)))
```

## 3.1 Explore Samples Table

View and explore the samples tibble created in the previous step.

```{r explore-samples}
cat("Samples tibble shape:", nrow(samples_df), "rows ×", ncol(samples_df), "columns\n\n")

if ("type" %in% colnames(samples_df)) {
  cat("Sample types distribution:\n")
  print(table(samples_df$type))
}

cat("\nFirst few rows:\n")
head(samples_df)
```

## 3.2 Explore Traits Table

View and explore the traits tibble.

```{r explore-traits}
cat("Traits tibble shape:", nrow(traits_df), "rows ×", ncol(traits_df), "columns\n\n")

if ("type" %in% colnames(traits_df)) {
  cat("Trait types distribution (top 10):\n")
  trait_counts <- sort(table(traits_df$type), decreasing = TRUE)
  print(head(trait_counts, 10))
}

cat("\nFirst few rows:\n")
head(traits_df)
```

## 3.3 Explore Experiments Table

View and explore the processed experiments tibble with polynomial fit data.

```{r explore-experiments}
cat("Experiments tibble shape:", nrow(experiments_df), "rows ×", ncol(experiments_df), "columns\n\n")

if ("r_squared" %in% colnames(experiments_df)) {
  cat("Fit quality statistics:\n")
  cat(sprintf("  Average R²: %.4f\n", mean(experiments_df$r_squared, na.rm = TRUE)))
  cat(sprintf("  Min R²: %.4f\n", min(experiments_df$r_squared, na.rm = TRUE)))
  cat(sprintf("  Max R²: %.4f\n", max(experiments_df$r_squared, na.rm = TRUE)))
}

if ("fracture_detected" %in% colnames(experiments_df)) {
  cat("\nFracture detection:\n")
  cat(sprintf("  Fractures detected: %d / %d\n",
              sum(experiments_df$fracture_detected, na.rm = TRUE),
              nrow(experiments_df)))
}

cat("\nFirst few rows:\n")
head(experiments_df)
```

------------------------------------------------------------------------

# Section 4: Analyze Outliers

Run the chunk below to perform hierarchical outlier analysis on the processed experimental data.

The analysis uses sigma-based detection (1σ, 2σ, 3σ) grouped by Family \> Species \> Subsample Type.

Configuration: - **Outlier trait threshold**: Percentage of traits that must be outliers to flag an experiment - **Sigma level**: Standard deviations threshold for outlier detection - Results saved to `processed_data/outlier_analysis.json`, `outlier_experiments.csv`, and `outlier_analysis.log`

```{r analyze-outliers, eval=FALSE}
# Source and run the outlier analysis script
source("analyse_outliers.R")
```

------------------------------------------------------------------------

# Next Steps: Custom Analysis

Now that you have the three tibbles loaded (`samples_df`, `traits_df`, `experiments_df`), you can perform custom analysis and create visualizations using the tidyverse and ggplot2.

**Examples**: - Explore relationships between traits and mechanical properties - Create scatter plots comparing different measurements - Filter data by family or species - Analyze trends across different sample types

Use the chunks below to write your own analysis code!

```{r load-viz-libraries}
# Load visualization libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Print columns of traits
cat("\nTraits DataFrame columns:\n")
print(colnames(traits_df))
```

## Example 1: Histogram of Modulus Traits

Distribution of modulus measurements across all samples.

```{r viz-modulus-histogram}
# Filter traits of type 'modulus'
if ("type" %in% colnames(traits_df) && "measurement" %in% colnames(traits_df)) {
  modulus_traits <- traits_df %>% filter(type == "modulus")

  ggplot(modulus_traits, aes(x = measurement)) +
    geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
    geom_density(aes(y = after_stat(count)), color = "darkblue", linewidth = 1) +
    labs(
      title = "Distribution of Modulus Traits",
      x = "Modulus Measurement",
      y = "Frequency"
    ) +
    theme_minimal()
}
```

## Example 2: Violin Plot for Pholcus phalangioides Dragline Samples

Comparison of trait measurements across different trait types for *Pholcus phalangioides* dragline silk.

```{r viz-pholcus-dragline}
# Filter for Pholcus phalangioides dragline samples
if ("sample.nomenclature" %in% colnames(traits_df) &&
    "sample.subsampletype" %in% colnames(traits_df)) {

  pholcus_dragline <- traits_df %>%
    filter(sample.nomenclature == "Pholcus phalangioides",
           sample.subsampletype == "dragline")

  ggplot(pholcus_dragline, aes(x = type, y = measurement)) +
    geom_violin(fill = "coral", alpha = 0.7) +
    scale_y_log10() +
    labs(
      title = "Trait Measurements for Pholcus phalangioides Dragline Samples",
      x = "Trait Type",
      y = "Measurement (log scale)"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Example 3: Strain at Break Across Pholcidae Species

Violin plot comparing strain at break measurements for different species within the Pholcidae family, grouped by subsample type.

```{r viz-pholcidae-strain}
# Filter for Pholcidae family and strainAtBreak trait
if ("sample.family" %in% colnames(traits_df)) {

  pholcidae_strain <- traits_df %>%
    filter(sample.family == "Pholcidae",
           type == "strainAtBreak")

  ggplot(pholcidae_strain,
         aes(x = sample.nomenclature, y = measurement,
             fill = sample.subsampletype)) +
    geom_violin(position = position_dodge(width = 0.9), alpha = 0.7) +
    labs(
      title = "Strain At Break Measurements for Pholcidae Species",
      x = "Species",
      y = "Strain At Break Measurement",
      fill = "Subsample Type"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_brewer(palette = "Set2")
}
```

## Example 4: Statistical Summary Table for Pholcidae

Summary statistics of measurements grouped by species, trait type, and subsample type for the Pholcidae family.

```{r table-pholcidae-summary}
# Create statistical summary table for Pholcidae family
if ("sample.family" %in% colnames(traits_df)) {

  pholcidae_summary <- traits_df %>%
    filter(sample.family == "Pholcidae") %>%
    group_by(sample.nomenclature, type, sample.subsampletype) %>%
    summarise(
      mean = mean(measurement, na.rm = TRUE),
      std = sd(measurement, na.rm = TRUE),
      min = min(measurement, na.rm = TRUE),
      max = max(measurement, na.rm = TRUE),
      .groups = "drop"
    )

  cat("\nStatistical Summary for Pholcidae Family:\n")
  print(pholcidae_summary, n = 50)  # Show first 50 rows

  # Create a nicely formatted table using knitr::kable
  library(knitr)
  kable(head(pholcidae_summary, 20),
        digits = 3,
        caption = "Statistical Summary of Measurements for Pholcidae Species (First 20 rows)")
}
```

------------------------------------------------------------------------

# Session Information

```{r session-info}
sessionInfo()
```
